### Inspiration for exercise
- https://www.andrew.cmu.edu/user/achoulde/94842/homework/regression_diagnostics.html
- You may find this useful to solve the task.

a) Linearity

Meaning: The relationship between predictors x and response y is linear.

Diagnosis: Plot residuals vs. fitted values (should show no pattern). Partial regression plots also help.

b) Homoscedasticity (constant variance)

Meaning: The variance of residuals is the same across all fitted values.

Diagnosis: Plot residuals vs. fitted values. If spread increases or decreases (a funnel shape), heteroscedasticity is present. Use statistical tests like Breusch–Pagan or White’s test.

c) Independence

Meaning: Observations (and residuals) are independent from each other.

Diagnosis: For time series, check autocorrelation plots (ACF) or Durbin–Watson test. For cross-sectional data, independence is usually assumed based on design.

d) Normality

Meaning: Residuals are normally distributed (important for inference, not prediction).

Diagnosis: Histogram or Q–Q plot of residuals. Formal tests like Shapiro–Wilk or Kolmogorov–Smirnov can be used.



```{r}
library(ggplot2)

set.seed(42)
n <- 1000
x <- 1:n

# Changeable parameters
# - Change the parameters to affect the generated data points below.
# - You may copy this code multiple times to answer all the questions in the exercise.
# - You may find it reasonable to argue for multiple violations from a single generated set of data points.

contant <- 0
trend <- 0
curve_magnitue <- 0
curve_period <- 100
curve_shift <- 0
normal_noise_magnitue <- 1
norm_noise_periode <- 10000
shift_norm_noise <- 500
non_normal_noise_magnitue <- 0
non_norm_noice_periode <- 10000
shift_non_norm_noise <- 500

# put y.gen into a function
y_gen <- function(
    contant, trend, curve_magnitue, curve_period,
    curve_shift, normal_noise_magnitue, norm_noise_periode,
    shift_norm_noise, non_normal_noise_magnitue,
    non_norm_noice_periode, shift_non_norm_noise
    ){
  contant + trend * x +
  curve_magnitue * sin((x/curve_period + curve_shift)*pi) + 
  normal_noise_magnitue*cos((
    x / norm_noise_periode + shift_norm_noise/norm_noise_periode)*pi
    )*rnorm(n, sd = 3) +
  non_normal_noise_magnitue*cos(
    (x/non_norm_noice_periode + shift_non_norm_noise/non_norm_noice_periode)*pi
  ) * rexp(n, rate = 0.2)
  }
```

```{r}
y.gen <- y_gen(contant, trend, curve_magnitue, curve_period,
               curve_shift, normal_noise_magnitue, norm_noise_periode,
               shift_norm_noise, non_normal_noise_magnitue,
               non_norm_noice_periode, shift_non_norm_noise)

p <- qplot(x, y.gen, ylab = "y") +
  geom_point(size = 0.1) +
  labs(title = "Data generate for linear regrestion")

# Display the plot
print(p)
```

```{r}
# ---- A: All assumptions hold ----
set.seed(1)

contant <- 0
trend <- 0.2                # linear trend
curve_magnitue <- 0           # no curvature
curve_period <- 100
curve_shift <- 0
normal_noise_magnitue <- 1
norm_noise_periode <- 1e9     # ~constant variance (cos ~ 1)
shift_norm_noise <- 0
non_normal_noise_magnitue <- 0
non_norm_noice_periode <- 1e9
shift_non_norm_noise <- 0

y.gen <- y_gen(
  contant=contant, trend=trend, curve_magnitue=curve_magnitue, curve_period=curve_period,
  curve_shift=curve_shift, normal_noise_magnitue=normal_noise_magnitue, norm_noise_periode=norm_noise_periode,
  shift_norm_noise=shift_norm_noise, non_normal_noise_magnitue=non_normal_noise_magnitue,
  non_norm_noice_periode=non_norm_noice_periode, shift_non_norm_noise=shift_non_norm_noise
)

qplot(x, y.gen) + geom_point(size=0.1) + labs(title="A) Data (assumptions hold)", y="y")
lm.gen <- lm(y.gen ~ x)
plot(lm.gen, which = 1)  # Residuals vs Fitted (should be patternless)
plot(lm.gen, which = 2)  # Q-Q (close to line)
plot(lm.gen, which = 3)  # Scale-Location (flat-ish)
plot(lm.gen, which = 5)  # Leverage (no standout points)
```
```{r}
# ---- B: Heteroscedasticity ----
set.seed(2)

contant <- 0
trend <- 0.01
curve_magnitue <- 0
curve_period <- 100
curve_shift <- 0
normal_noise_magnitue <- 1
norm_noise_periode <- 500      # variance oscillates strongly across x
shift_norm_noise <- 0
non_normal_noise_magnitue <- 0
non_norm_noice_periode <- 1e9
shift_non_norm_noise <- 0

y.gen <- y_gen(
  contant=contant, trend=trend, curve_magnitue=curve_magnitue, curve_period=curve_period,
  curve_shift=curve_shift, normal_noise_magnitue=normal_noise_magnitue, norm_noise_periode=norm_noise_periode,
  shift_norm_noise=shift_norm_noise, non_normal_noise_magnitue=non_normal_noise_magnitue,
  non_norm_noice_periode=non_norm_noice_periode, shift_non_norm_noise=shift_non_norm_noise
)

qplot(x, y.gen) + geom_point(size=0.1) + labs(title="B) Data (heteroscedastic)", y="y")
lm.gen <- lm(y.gen ~ x)
plot(lm.gen, which = 1)  # Residuals vs Fitted: changing spread (violates homoscedasticity)
plot(lm.gen, which = 2)  # Q-Q
plot(lm.gen, which = 3)  # Scale-Location: upward/downward pattern (non-constant variance)
plot(lm.gen, which = 5)  # Leverage


```
```{r}
# ---- C: Nonlinearity ----
set.seed(3)

contant <- 0
trend <- 0.0                  # keep linear part minimal to expose curvature
curve_magnitue <- 20          # visible curvature
curve_period <- 100           # periodic curve in mean
curve_shift <- 0
normal_noise_magnitue <- 1
norm_noise_periode <- 1e9     # ~constant variance to isolate nonlinearity
shift_norm_noise <- 0
non_normal_noise_magnitue <- 0
non_norm_noice_periode <- 1e9
shift_non_norm_noise <- 0

y.gen <- y_gen(
  contant=contant, trend=trend, curve_magnitue=curve_magnitue, curve_period=curve_period,
  curve_shift=curve_shift, normal_noise_magnitue=normal_noise_magnitue, norm_noise_periode=norm_noise_periode,
  shift_norm_noise=shift_norm_noise, non_normal_noise_magnitue=non_normal_noise_magnitue,
  non_norm_noice_periode=non_norm_noice_periode, shift_non_norm_noise=shift_non_norm_noise
)

qplot(x, y.gen) + geom_point(size=0.1) + labs(title="C) Data (nonlinear mean)", y="y")
lm.gen <- lm(y.gen ~ x)
plot(lm.gen, which = 1)  # Residuals vs Fitted: curved pattern (violates linearity)
plot(lm.gen, which = 2)  # Q-Q
plot(lm.gen, which = 3)  # Scale-Location
plot(lm.gen, which = 5)  # Leverage

```


```{r}
# ---- D: Non-normal residuals ----
set.seed(4)

contant <- 0
trend <- 0.01
curve_magnitue <- 0
curve_period <- 100
curve_shift <- 0
normal_noise_magnitue <- 0     # turn off normal part to isolate non-normality
norm_noise_periode <- 1e9
shift_norm_noise <- 0
non_normal_noise_magnitue <- 1 # exponential noise component on
non_norm_noice_periode <- 1e9  # ~constant scale
shift_non_norm_noise <- 0

y.gen <- y_gen(
  contant=contant, trend=trend, curve_magnitue=curve_magnitue, curve_period=curve_period,
  curve_shift=curve_shift, normal_noise_magnitue=normal_noise_magnitue, norm_noise_periode=norm_noise_periode,
  shift_norm_noise=shift_norm_noise, non_normal_noise_magnitue=non_normal_noise_magnitue,
  non_norm_noice_periode=non_norm_noice_periode, shift_non_norm_noise=shift_non_norm_noise
)

qplot(x, y.gen) + geom_point(size=0.1) + labs(title="D) Data (non-normal errors)", y="y")
lm.gen <- lm(y.gen ~ x)
plot(lm.gen, which = 1)  # Residuals vs Fitted
plot(lm.gen, which = 2)  # Q-Q: systematic deviations (skew/heavy tails)
plot(lm.gen, which = 3)  # Scale-Location
plot(lm.gen, which = 5)  # Leverage

```
a) Linearity violated

What it means: The conditional mean 
𝐸
[
𝑌
∣
𝑋
]
E[Y∣X] isn’t a straight line in 
𝑋
X.

In your generator: A curved mean like a sinusoid breaks linearity (e.g., curve_magnitue > 0 so 
𝑌
=
𝛽
0
+
𝛽
1
𝑋
+
𝐴
sin
⁡
(
⋅
)
+
𝜀
Y=β
0
	​

+β
1
	​

X+Asin(⋅)+ε).

What you’ll see: Residuals vs Fitted shows a systematic curve/wave (not a random cloud around 0). The linear fit is biased because it tries to force a line through a curved signal.

b) Homoscedasticity violated

What it means: The error variance changes with 
𝑋
X (heteroscedasticity).

In your generator: Make the noise scale depend on 
𝑋
X (e.g., your cosine envelope via norm_noise_periode/shift_norm_noise so the spread oscillates with 
𝑥
x).

What you’ll see: Residuals vs Fitted (and Scale–Location) show changing spread—funnel shapes or periodic widening/narrowing. Coefficient SEs become unreliable unless you use robust SEs or model the variance.

c) Independence violated

What it means: Errors are correlated across observations (common with time index 
𝑥
=
1
:
𝑛
x=1:n).

In your data: If errors contain persistence/seasonality (e.g., any serial pattern in the noise term), residuals at nearby 
𝑥
x are autocorrelated.

What you’ll see: Residuals vs Fitted may show runs; an ACF of residuals has significant lags; Durbin–Watson deviates from ~2. Consequence: SEs are understated and tests overstate significance. (OLS estimates remain unbiased if regressors are exogenous but are inefficient.)

d) Normality violated

What it means: Error distribution isn’t normal (e.g., skewed or heavy‐tailed).

In your generator: Turn on the exponential component (non_normal_noise_magnitue > 0) or any heavy‐tailed noise.

What you’ll see: Q–Q plot deviates from the diagonal: bowed tails (heavy tails), or systematic curvature (skew). Histograms show skew/kurtosis. OLS coefficients are still unbiased/consistent, but t/F inference can be off for small samples (robust or bootstrap SEs help).



4)
a)
Correlation is a standardized measure of the linear relationship between two random variables. It tells us how strongly and in what direction (positive or negative) two variables move together, on a scale from –1 to +1. Covariance measures the joint variability of two random variables: A positive covariance means that when X is above its mean, Y tends to be above its mean as well (and vice versa).

b)

Cov(αX,Y)=αCov(X,Y)

c)

Correlation is unit-free — rescaling doesn’t matter, but flipping the axis flips the sign.
If you multiply a variable by a positive number (e.g. converting height from meters to centimeters), you’re just stretching or shrinking the scale. The relationship between X and Y doesn’t change at all. So the correlation stays exactly the same.

Correlation cares only about the relationship in variation, not about absolute levels.
Adding or subtracting a constant is just moving the scale up or down (e.g. changing the temperature scale from Celsius to Kelvin, or shifting all salaries up by 10,000).

This shift doesn’t change how the variables move together. The “pattern of ups and downs” between X and Y is unchanged.

d)

Partial correlation measures the linear link between 
X and Y after removing the linear influence of Z from both. Equivalently: regress X on Z, regress Y on Z, and then correlate the two sets of residuals.

Numerator: start with the raw correlation between X and Y; subtract the part that can be “explained” by both being related to Z.
Denominator: rescales by how much unique variation remains in X and Y after accounting for Z.
